---
title: "pixelate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pixelate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pixelate)
```

## Before you start

This package was designed to pixelate point estimates generated by a geostatistical model that also generates measures of uncertainty i.e. credible or confidence intervals. 

We illustrate this using the median posterior predictive rate of *Plasmodium falciparum* in 2-10 year olds globally, 2000-2017. 

Tip: on the Malaria Atlas Project website (https://map.ox.ac.uk/) you can "clip a surface to an area, either by selecting a region, or by uploading your own custom shapefile"

We could store this in inst/extdata, but it's very very large. Instead, we suggest you download and save files locally, then proceed to step 2. 

# Step one: format point estimates and credible/confidence intervals
